---
title: "Deriving exposures with gaiaCore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deriving exposures with gaiaCore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

This vignette follows the Gaia pipeline from a catalog entry to exposure rows in an OMOP CDM. It needs a running gaiaDB
(see [gaiaDocker](https://github.com/OHDSI/gaiaDocker)), so the code is shown but not run.

## 1. Connect

```{r}
library(gaiaCore)
connectionDetails <- createGaiaConnectionDetails(server = "gaia-db/gaiacore", password = Sys.getenv("GAIA_POSTGRES_PASSWORD"))
connection <- connectGaia(connectionDetails)
```

## 2. Ingest the sources

A dataset is ingested from its gaiaCatalog entry: its metadata is registered, the data are downloaded and loaded into PostGIS, and the
geometry is cleaned. Ingest boundaries before datasets that are joined to them.

```{r}
ingestDatasource(connection, "us_2023_county_tl")
ingestDatasource(connection, "us_2014_2019_monthly_pm25_by_county_cdc")
listDatasources(connection)
listVariables(connection)
```

`loadVariables()` then builds the geometry and attribute tables of every variable of a dataset. The county PM2.5 dataset has four
variables (maximum, median, mean, population-weighted mean); the next step joins only the one you choose.

```{r}
loadVariables(connection, "us_2014_2019_monthly_pm25_by_county_cdc", geomLabel = "name", variableNodata = -999)
```

## 3. Locations

Locations come from an OMOP CDM with the Gaia extension tables or from data frames.

```{r}
loadLocationsFromOmop(connection, cdmSchema = "omopgis")
# or
loadLocations(connection, location, locationHistory)
validateLocations(connection)
```

## 4. The spatial-temporal join

```{r}
spatialJoin(connection, "pm25_mean_pred", "us_2014_2019_monthly_pm25_by_county_cdc",
            spatialOperator = "ST_Within", exposureTypeConceptId = 2052499878)
summarizeExposure(connection)
```

A person who moves in the middle of a month has two partial-month rows for that month: the monthly value is clipped to each residence interval.

## 5. Check and copy into the CDM

```{r}
checkExposure(connection, expectedUnit = NULL, valueRange = c(0, 200))
copyExposureToOmop(connection, cdmSchema = "omopgis")
```

## 6. Analyse

Exposure over a window that does not line up with the monthly rows, such as a pregnancy:

```{r}
exposure <- getExposure(connection, conceptId = 2052499839)
dayWeightedExposure(exposure, windows)
```

An exposure effect with a confidence interval that holds up with few, unbalanced clusters:

```{r}
fitExposureEffect(copd ~ pm25 + ses + age + female, data, exposure = "pm25", cluster = "county")
```

```{r}
disconnectGaia(connection)
```
